---
title: "Analytical_Depletion_Example"
author: "Christopher Dory"
date: "2025-11-05"
output: github_document
---

```{r setup, include=FALSE,  message = FALSE, error = FALSE, warning = FALSE}
## IMPORTANT 
## POINT THIS TO YOUR DIRECTORY
knitr::opts_knit$set(echo = TRUE,
                      root.dir = getwd())
```

```{r setup2, include=FALSE, message = FALSE, error = FALSE, warning = FALSE}
#===========================================================================================
# Requirements Load
#===========================================================================================
library(sf)
library(sp)
library(raster)
library(colorspace)
library(terra)
library(lubridate)
library(stringr)
library(here)
library(DT)
library(data.table)
library(devtools)
options(scipen = 100)
```




```{r sourcefiles, echo = FALSE, warning = FALSE, message = FALSE}
source_url('https://raw.githubusercontent.com/Dory111/Analytical-Depletion-Function/refs/heads/main/General_Stream_Depletion_Function.R')
```

# Introduction

This will be an example of how to calculate stream depletions using the fastest method <br/>
from Zipper et al. (https://doi.org/10.1029/2018WR024403). The fastest method uses the 'local area' proximity criteria, the 'inverse distance' apportionment method, and the 'glover'(glover-balmer) analytical model. <br/>
<br/>
For this we will first load in the following files: <br/>
Shasta groundwater basin <br/>
Shasta streams <br/>
<br/>
And then assign an 'influence radius' to a series of randomly generated wells in the domain <br/>
The influence radius describes the radius over which a well is expected to produce meaningful <br/>
depletion. It is defined by Zipper et al. (2019) to be twice the maximum distance from <br/>
any land point within the domain to its closest stream segment. <br/>
This ensures that each well has 1 >= segment associated with it. <br/> 
<br/>
<br/>
**NOTES** <br/>
1) Though the program will attempt to correct differing projections, it is best if they <br/>
&nbsp;&nbsp;&nbsp; are projected beforehand.<br/>
2) To avoid depletions predicted for inappropriate locations, please pre-process streams <br/>
&nbsp;&nbsp;&nbsp; so that they are only within the groundwater basin <br/>

```{r loadshapefiles, warning = FALSE, message = FALSE}
Shasta_Streams <- st_read(file.path(here::here(),
                                    'Data',
                                    'Shapefiles',
                                    'shasta_mainstreams.shp'), quiet = TRUE)
Shasta_Streams <- st_zm(Shasta_Streams)
Shasta_B118 <- st_read(file.path(here::here(),
                                 'Data',
                                 'Shapefiles',
                                 'Bulletin 118 Boundary_shasta.shp'), quiet = TRUE)
Shasta_B118 <- st_transform(Shasta_B118, 3310)
Shasta_Streams <- st_intersection(Shasta_Streams, Shasta_B118$geometry)
```


```{r createrandomwells, warning = FALSE, message = FALSE}
bb <- st_bbox(Shasta_B118)
set.seed(1)
x <- runif(n = 300, min = bb['xmin'], max = bb['xmax'])
y <- runif(n = 300, min = bb['ymin'], max = bb['ymax'])
random_wells <- cbind(x,y) %>% as.data.frame()
random_wells <- st_as_sf(random_wells,
                         coords = c('x','y'),
                         crs = st_crs(Shasta_B118),
                         na.fail = FALSE)
random_wells <- st_intersection(random_wells,
                                Shasta_B118$geometry)
```

```{r randomwellplot, warning = FALSE, message = FALSE, fig.height = 10, fig.width = 8, echo = FALSE}
plot(st_geometry(Shasta_B118), lwd = 3)
plot(st_geometry(Shasta_Streams), col = 'dodgerblue4', lwd = 2, add = TRUE)
plot(st_geometry(random_wells), add = TRUE, pch = 19, col = '#A21')
legend(x = 'topleft',
       lwd = c(3,2,NA),
       pch = c(NA,NA,19),
       col = c('black','dodgerblue4','#A21'),
       legend = c('Shasta Groundwater Basin','Main Streams','Randomly Placed Wells'))
```

<br/>
<br/>
<br/>
<br/>
<br/>
<br/>

# Local Area Criteria

To implement Zippers 'Local Area' criteria we will create a 1km grid over the domain, <br/>
take the centroids of those grid cells, clip them by the groundwater basin <br/>
calculate the distance of each centroid to its closest stream, and then <br/>
take double the maximum<br/>
<br/>
1km was chosen due to computing power limitations<br/>
Grid size will change results slightly, but by increasing the number of grid<br/>
cells by 100 the result only changes by about 1%<br/>

```{r localareacriteria, warning = FALSE, message = FALSE}
#-------------------------------------------------------------------------------
# make grid and centroids
grid <- st_make_grid(Shasta_B118,
                     cellsize = 1000)

centroids <- st_centroid(grid)
centroids <- st_intersection(centroids, Shasta_B118$geometry)
#-------------------------------------------------------------------------------

#-------------------------------------------------------------------------------
# find influence radius
dist_matrix <- st_distance(centroids, Shasta_Streams)
diag(dist_matrix) <- NA
min_distances <- apply(dist_matrix, 1, min, na.rm = TRUE)

influence_radius <- max(min_distances)*2
#-------------------------------------------------------------------------------
```
<br/>
<br/>
For Shasta the influence radius is calculated <br/>
here as `r round(influence_radius/1000,0)` kilometers. <br/>
The below visualization shows wells, with their sized scaled by the length of river <br/>
that they effect relative to the maximum length. <br/>
As expected, those at the periphery effect less reach. <br/>

```{r localplot, warning = FALSE, message = FALSE, fig.height = 10, fig.width = 8, echo = FALSE}
l <- list()
for(i in 1:nrow(random_wells)){
  b <- st_buffer(random_wells[i, ], influence_radius)
  int <- st_intersection(Shasta_Streams, b$geometry)
  l[[i]] <- sum(st_length(int))
}
l <- unlist(l)
random_wells$length <- l
random_wells$length_eff <- random_wells$length/max(random_wells$length)
random_wells$length_eff[random_wells$length_eff < 0.1] <- 0.1


bb <- st_bbox(Shasta_B118)
plot(st_geometry(Shasta_B118), lwd = 3,
     main = paste0('min L (km): ',
                   round(min(random_wells$length)/1000,0),
                   '    ',
                   'max L (km)',
                   round(max(random_wells$length)/1000,0)),
     cex.main = 0.8)
plot(st_geometry(Shasta_Streams), col = 'dodgerblue4', lwd = 2, add = TRUE)
plot(st_geometry(random_wells), add = TRUE, pch = 19, col = '#A21', cex = 2*random_wells$length_eff)
p <- st_point(c((bb['xmax'] + bb['xmin'])/2,
                (bb['ymax'] + bb['ymin'])/2)) %>% st_sfc(crs = st_crs(Shasta_B118))
plot(st_buffer(p, influence_radius), add = TRUE, lty = 2, border = 'black')
legend(x = 'topleft',
       lwd = c(3,2,NA,2),
       lty = c(1,1,NA,2),
       pch = c(NA,NA,19,NA),
       col = c('black','dodgerblue4','#A21','black'),
       legend = c('Shasta Groundwater Basin',
                  'Main Streams',
                  'Randomly Placed Wells',
                  'Example Influence Radius'))
```
<br/>
The influence radius calculated here is nearly equivalent to the 'whole domain' proximity criteria, where each well is considered to have some, even if vanishingly small, potential to effect each stream. In the study area of Zipper et al. (2019), the Navarro River Watershed, streams are more evenly distributed than they are in the Shasta Valley, yielding a smaller (and more reasonable) influence radius.
<br/>
<br/>
However, as we will see in the following section, just because a stream is within the influence radius of a well does not mean a notable amount of depletion are apportioned to it. Proximate streams will always be prioritized over farther ones, and its likely that far reaches will have << 1% of the wells pumping assigned to them.
<br/>
<br/>
Furthermore, analytical solutions such as the Glover-Balmer solution (1954) take distance as an argument. And so even in the extreme case that there is one stream and one well separated by an inordinate distance, say 100km, while all of the wells pumping will be assigned to the stream the distance is such that the depletion fraction will never be noticeable.
<br/>
<br/>
This is not to say that the proximity criteria arguments are moot, as small errors can compound if there are many streams and many wells each with slight mis-assignments, but rather that the variance in model scenarios is likely to be the volume of depletion on an individual reach and not the identification of the most impacted reach. This conclusion is corroborated by figure 4 of Zipper et al. (2019).
<br/>
<br/>
<br/>
<br/>
<br/>
<br/>

# Depletion Apportionment Criteria

For the depletion apportionment criteria here we will use inverse distance. This is defined by the following equation:<br/>
$$\Large f_{(r,w)} = \frac{\frac{1}{d_{(r,w)}^W}}{\sum_{j=1,n} \frac{1}{d_{(r_{j},w)}^W}}$$
<br/>
&nbsp;&nbsp;&nbsp;&nbsp;Where:<br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $f_{(r,w)} [-]=$ the fraction of the depletion potential from well 'w' assigned to reach 'r'<br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $d_{(r,w)} [L]=$ the distance of well 'w' to reach 'r'<br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $W =[-]$ the power of the equation (1 for inverse distance, 2 for inverse distance squared)<br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $\sum_{j = 1,n} \frac{1}{d_{(r_{j},w)}^W} [L]=$ the sum of the distances from well 'w' to all reaches (j = 1,n) assigned to it<br/><br/>

### Example

<br/>
We have a well 'w' that is connected to four reaches, 'A','B','C' that are distances 10,40,50 from the well. Using the above formula we would arrive at the following for reach 'A'.<br/>
$$\Large f_{(A,w)} = \frac{0.1}{0.1 + 0.025 + 0.02} = 0.689$$
<br/>
So we see that the closest reach 'A' is apportioned the lions share of whatever depletions are calculated from well 'w'. No matter what depletion apportionment we use, the sum should always equal to 1.
<br/>
<br/>
<br/>
<br/>
<br/>
<br/>

# Depletion Analytical Model

This example will, as mentioned, use the Glover-Balmer solution of Glover and Balmer (1954)  https://doi.org/10.1029/TR035i003p00468.<br/>
Their equation is based on the earlier work of Theis in potentiometric surface drawdown from groundwater pumping, and assumes (in contrast to later, more computationally demanding methods) a linear stream that fully penetrates the aquifer with no streambed clogging layer. Their analytical solution is then multiplied by the fraction $f_{(r,w)}$ derived above.

$$\Large Q_{(r,w)} = Q_{w}*\text{erfc} \LARGE (\Large\sqrt{\frac{Sd_{(r,w)}^2}{4Tt}}\LARGE) \Large * f_{(r,w)}$$

<br/>
&nbsp;&nbsp;&nbsp;&nbsp;Where:<br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $Q_{(r,w)} [\frac{L^3}{T}]=$ the cumulative depletion experienced by reach r as a result of well w<br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $Q_{w} [\frac{L^3}{T}]=$ the pumping rate of well w<br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $\text{erfc} [-]=$ the complimentary error function, as the value inside approaches 0 it will return 1<br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $S [\frac{L}{L}]=$ the storativity of the aquifer<br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $d_{(r,w)}^2 [L]=$ the distance between the reach and the well squared<br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $T [\frac{L^2}{T}]=$ the transmissivity of the aquifer, described also as the thickness of the aquifer multiplied by its conductivity $(bK)$ <br/>
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp; $t [T]=$ the time at which the equation is to be evaluated<br/>


<font size = '4'>**Important**</font><br/>
While technically any units can be used, if the spatial projection of the data is one typically used (such as NAD83) the length units when distances are calculated will be in meters. Therefore the user should convert all units into metric beforehand (such as gallons per day to cubic meters per day).
<br/>
<br/>

### Example
For our example we'll just assign some arbitrary numbers for storativity, transmissivity, and pumping. The exact values aren't important at this moment, but demonstrating the correct format for them is.

```{r assignvals, warning = FALSE}
# giving each well some random aquifer parameters
random_wells <- random_wells[,-c(2,3)]

random_wells$Stor <- runif(n = nrow(random_wells),
                           min = 0.02, max = 0.1)
random_wells$Tr <- runif(n = nrow(random_wells),
                         min = 50, max = 200)


# assign pumping to each well
# where each row is a well, each column is a timestep, and each value
# in a row, column pair is the pumping rate at that timestep for that well
pumping <- matrix(runif(n = nrow(random_wells),
                        min = 100,
                        max = 300),
                  nrow = nrow(random_wells),
                  ncol = 250,byrow = TRUE)
pumping <- cbind(pumping, matrix(0,
                                 nrow = nrow(random_wells),
                                 ncol = 115))
pumping <- cbind(pumping, matrix(runif(n = nrow(random_wells),
                                       min = 100,
                                       max = 300),
                                 nrow = nrow(random_wells),
                                 ncol = 115,byrow = TRUE))
pumping <- cbind(pumping, matrix(0,
                                 nrow = nrow(random_wells),
                                 ncol = 115))
```
<br/>
<br/>
**Pumping Structure**<br/>
Each row is a well<br/>
Each column is a timestep<br/>
Each row,column combination contains a pumping rate<br/>
<br/>
If the timestep is days then each row,column combination will have the number of $L^3$ pumped in that day.<br/>
For visibility only the first 5 columns are shown here.
```{r table1, echo = FALSE, fig.width = 10, fig.height = 8}
data.table(pumping[,c(1:5)])
```
<br/>
<br/>
**Well Structure**<br/>
Wells have a column for aquifer properties.
```{r table2, echo = FALSE}
data.table(random_wells)
```


```{r examplecalc, warning = FALSE, fig.height = 10, fig.width = 8, message = FALSE, error = FALSE}
calculate_stream_depletions(streams = Shasta_Streams,
                            wells = random_wells,
                            pumping = pumping,
                            influence_radius = influence_radius, # influence radius calculated above
                            proximity_criteria = 'local area',
                            apportionment_criteria = 'inverse distance',
                            analytical_model = 'glover',
                            data_out_dir = file.path(getwd(),'Output'),
                            diag_out_dir = file.path(getwd(),'Output'),
                            stor_coef_key = 'Stor', # where to find storage coef in well set
                            well_transmissivity_key = 'Tr') # where to find transmissivity in well set
```
<br/>
<br/>
<br/>
<br/>
<br/>
<br/>

# Results

<br/>
In our example we can look to a random reach to see how our intermittent pumping effected depletions.<br/>
<br/>
Of note is that when pumping is at 0 depletions are still accumulating from pumping, indicating both a delayed response (as expected) from pumping and that the break from pumping was not long enough to reduce depletions.<br/>
```{r finalexampleplot, warning = FALSE, fig.height = 10, fig.width = 8, echo = FALSE}
dep <- read.csv(file.path(getwd(),'Output','volumetric_depletions_by_reach.csv'))
nf <- layout(matrix(c(1,2),
                    ncol = 1,
                    nrow = 2), heights = c(0.5,1))
plot(pumping[1,],
     ylab = 'L^3/T',
     cex.lab = 1.4)
rect(par('usr')[1],
     par('usr')[3],
     par('usr')[2],
     par('usr')[4],
     col = 'gray97')
grid()
points(pumping[1,],
       cex = 1,
       col = 'dodgerblue4',
       pch = 19)
box()

x_range <- par('usr')[2] - par('usr')[1]
y_range <- par('usr')[4] - par('usr')[3]
text(xpd = NA,
     labels = 'Random Pumping with Periods of Zero Pumping',
     pos = 4,
     x = par('usr')[1] - x_range*0.01,
     y = par('usr')[4] + y_range*0.1,
     cex = 1.6)

plot(as.vector(unlist(dep[1,-c(1)])),
     xlab = 'Timestep',
     ylab = 'L^3/T',
     cex.lab = 1.4)
rect(par('usr')[1],
     par('usr')[3],
     par('usr')[2],
     par('usr')[4],
     col = 'gray97')
grid()
lines(as.vector(unlist(dep[1,-c(1)])),
      cex = 1,
      col = 'dodgerblue4',
      pch = 19,
      lwd = 3)
box()
x_range <- par('usr')[2] - par('usr')[1]
y_range <- par('usr')[4] - par('usr')[3]
text(xpd = NA,
     labels = 'Depletions on Reach 1',
     pos = 4,
     x = par('usr')[1] - x_range*0.01,
     y = par('usr')[4] + y_range*0.05,
     cex = 1.6)

```

<br/>
<br/>
<br/>
As aquifer properties are assigned randomly, we also expect a random assortment in the spatial variability of our depletions, with in general streams with a large number of proximal wells, or a small number of extremely proximal wells, affected the most. <br/>
<br/>
Due to the random nature of our well location generation, there are some that are in unrealistic locations on top of the streams themselves. However, the important thing here is demonstrating that the method works, as again the numbers are randomly assigned and come with no context as to whether the absolute volume of depletions are large in relation to the flows in the rivers.<br/>
<br/>
Here streams are colored by how much of the global maximum (`r round(max(dep, na.rm = T),0)` [ $\frac{L^3}{T}$ ]) depletions they experienced .
```{r finalexampleplot2, warning = FALSE, fig.height = 10, fig.width = 8, echo = FALSE}
mdep <- c()
for(i in 1:nrow(Shasta_Streams)){
  mdep <- append(mdep,
                 max(as.vector(unlist(dep[i,]))))
}
mdep <- mdep/max(mdep)
Shasta_Streams$mdep <- mdep


plot(st_geometry(Shasta_B118), lwd = 3)
plot(st_geometry(random_wells), add = TRUE, pch = 19, col = '#A21')
plot(st_geometry(Shasta_Streams[Shasta_Streams$mdep < 0.25,]),
     col = 'dodgerblue4', lwd = 3, add = TRUE)
plot(st_geometry(Shasta_Streams[Shasta_Streams$mdep < 0.5 & Shasta_Streams$mdep >= 0.25,]),
     col = 'green', lwd = 3, add = TRUE)
plot(st_geometry(Shasta_Streams[Shasta_Streams$mdep < 0.75 & Shasta_Streams$mdep >= 0.5,]),
     col = 'goldenrod', lwd = 3, add = TRUE)
plot(st_geometry(Shasta_Streams[Shasta_Streams$mdep >= 0.75,]),
     col = 'red', lwd = 3, add = TRUE)
legend(x = 'bottomleft',
       lwd = c(3,NA,NA,NA,NA,NA,NA),
       pch = c(NA,19,15,15,15,15,15),
       pt.cex = c(NA,1.6,1.8,1.8,1.8,1.8,1.8),
       col = c('black','#A21','white','dodgerblue4','green','goldenrod','red'),
       legend = c('Shasta Groundwater Basin',
                  'Randomly Placed Wells',
                  '% Of Maximum Depletion Potential',
                  '< 25%',
                  '25 - 50%',
                  '50 - 75 %',
                  '> 75 %'))
```